---
title: "Module 05: Summaries and Decisions"
subtitle: "Every summary hides a decision. Make it on purpose."
author: "Chris Chizinski"
lang: en
footer: "[NRES 800 Fall 2026 GitHub repository](https://github.com/chrischizinski/nres800-slides)"
format:
  revealjs:
    theme: ../styles/pp.scss
    css: ../styles/slide-extras.css
    transition: fade
    slide-number: true
    incremental: false
    embed-resources: true
---

```{r}
#| include: false
library(gt)
library(countdown)

# One table style for every slide table (instructor convention: gt tables).
slide_table <- function(df) {
  gt(df) |>
    fmt_markdown(columns = everything()) |>
    tab_options(
      table.font.size = px(24),
      column_labels.font.weight = "bold",
      table.width = pct(100),
      data_row.padding = px(6)
    ) |>
    opt_row_striping()
}
```

## Slide Accessibility Note
- **Images:** none; diagrams are HTML text with labels.
- **Speaker Notes:** These slides follow the Module 05 follow-along script.
  They name the purpose, vocabulary, challenges, and checklist; the live
  coding happens in `class/week_05_follow_along.R`.

<!-- visual-aid: week_05 -->
## Today's Workflow

::: {.diagram-flow .diagram-compact aria-label="Workflow from clean data to a checked answer: decide which rows count, decide categories with if_else and case_when, group_by, summarize, check against a known total"}
<div class="diagram-node">Clean data</div>
<div class="diagram-arrow" aria-hidden="true">-&gt;</div>
<div class="diagram-node">Decide which rows count</div>
<div class="diagram-arrow" aria-hidden="true">-&gt;</div>
<div class="diagram-node">Decide categories</div>
<div class="diagram-arrow" aria-hidden="true">-&gt;</div>
<div class="diagram-node"><code>group_by</code> + <code>summarize</code></div>
<div class="diagram-arrow" aria-hidden="true">-&gt;</div>
<div class="diagram-node">Check against a known total</div>
:::

<p class="diagram-caption">A total you cannot check is a guess.</p>

## Before We Start

- Open **this module's repo** (accepted in Classroom 50 before class).
- Open `week_05.Rproj`, then `class/week_05_follow_along.R`.
- Run `here::here()`: it should print your repo folder.
- Backup: the follow-along is also on this slide site.

<a class="script-download" href="../scripts/week_05_follow_along.R" download style="background-color: #d00000 !important; color: #fff !important;">Download the Module 05 follow-along script (.R)</a>

::: {.notes}
Anyone who could not clone: pair them with a neighbor for Cycle 1 and fix the clone during Challenge 1.
:::

## Recap: Module 04

Answer in your script first:

1. After `read_csv()`, what do you run first on real data, and why?
2. A filter dropped 39 of 109 rows. What must you report with the answer?
3. What does `1987 %/% 10 * 10` give?

## Learning Goals

By the end of today you can:

- summarize a table, and one row **per group**, with `group_by()` + `summarize()`
- use `count()` as the shortcut, including two-column counts
- handle `NA` in summaries: `na.rm = TRUE`, and **report how many were missing**
- make categories with `if_else()` (two) and `case_when()` (many)
- spot **total rows mixed with detail rows**, and summarize without double counting
- explain how the rows and the denominator change a percentage

## The Data: Great Lakes Commercial Catch

- Great Lakes Fishery Commission commercial fish production
- **65,706 rows x 7 columns**, 1867-2015, six lakes
- `values` = catch in **thousands of pounds**
- Used exactly as released. Via TidyTuesday (2021-06-08)

::: {.notes}
Same agency as last week's stocking data. Ask: what would a fisheries manager want summarized from this?
:::

## Vocabulary

```{r}
#| echo: false
tibble::tribble(
  ~Term, ~Meaning,
  "summary", "one number that stands for many rows: a count, total, mean",
  "grouping", "`group_by()` splits rows; `summarize()` returns one row per group",
  "`na.rm = TRUE`", "drop `NA` before computing: a decision to **report**, not a fix",
  "total row", "a row that already sums other rows; include it and you double count",
  "denominator", "what a percentage is \"out of\"; change it, change the answer"
) |>
  slide_table()
```

## Cycle 1: One Row Per Group

```r
fishing <- read_csv(here("class", "data", "fishing.csv")) |>
  clean_names()

fishing |> count(lake)                        # you know this one
fishing |> summarize(total = sum(values))     # NA. Why?

fishing |>
  group_by(lake) |>
  summarize(
    rows      = n(),
    n_missing = sum(is.na(values)),           # report what you dropped
    total     = sum(values, na.rm = TRUE)
  ) |>
  arrange(desc(total))
```

## Why Is The Sum `NA`?

- 21,916 of 65,706 `values` are `NA` (a third of the file)
- One unknown value makes the total unknown; R refuses to guess
- `na.rm = TRUE` drops them. That is **your** decision
- So report `n_missing` right beside the total, every time

. . .

`sum(is.na(values))`: `is.na()` gives TRUE/FALSE, `sum()` counts the TRUEs.

## A Number That Should Worry You

```r
fishing |>
  filter(lake == "Michigan", species == "Lake Whitefish", year == 2000) |>
  summarize(total = sum(values, na.rm = TRUE))
#> 14381
```

Is that the U.S. commercial lake whitefish catch from Lake Michigan in 2000?

## Challenge 1: Catch Audit (15 min)

**Question:** where did 14,381 come from?

- **Base:** show the rows behind the total (`region`, `values`). Compare
  with the `U.S. Total` row. How many times too big? Explain with
  `distinct(lake, region)`
- **Extension:** for every lake, all-rows sum vs U.S. total sum. Which lake
  is furthest off, and why is that ratio not only double counting?

```{r}
#| echo: false
countdown(minutes = 15)
```

## The Same Catch, Counted Three Times

```{r}
#| echo: false
tibble::tribble(
  ~region, ~values, ~`what it is`,
  "Green Bay (MI)", "1,158", "area",
  "Mich. Proper (MI)", "1,871", "area",
  "**MI State Total**", "**3,029**", "sum of the two above",
  "Green Bay (WI)", "116", "area",
  "Mich. Proper (WI)", "1,648", "area",
  "**WI State Total**", "**1,765**", "sum of the two above",
  "Illinois (IL), Indiana (IN)", "NA", "not reported",
  "**U.S. Total**", "**4,794**", "sum of the state totals"
) |>
  slide_table() |>
  cols_align(align = "right", columns = values)
```

Lake Michigan lake whitefish, 2000 (thousand lb). All rows sum to 14,381.

## Cycle 2: Categories You Decide

```r
fishing |>
  mutate(recorded = if_else(is.na(values), "missing", "recorded")) |>
  count(lake, recorded)

fishing <- fishing |>
  mutate(
    region_level = case_when(
      region %in% us_totals     ~ "US total",
      region %in% canada_totals ~ "Canada total",
      region %in% state_totals  ~ "state total",
      .default = "area"
    )
  )

fishing |> count(region_level, region)   # did every region land right?
```

## How `case_when()` Reads

- Each line: **if this ~ then that**
- Lines are checked **top to bottom**; the **first TRUE wins**
- `.default` catches every row no line matched (forget it: `NA`)
- Always `count()` the result to check it

. . .

```r
era = case_when(
  year < 2000 ~ "1950-1999",   # first: now 1867 lands here too
  year < 1900 ~ "1867-1899",   # never reached
  ...
)
```

## Now The Totals Are Right

```r
fishing |>
  filter(region_level == "US total") |>
  group_by(lake) |>
  summarize(n_missing = sum(is.na(values)),
            total = sum(values, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(total))
```

- Michigan first: 3,072,965 thousand lb
- Saint Clair last, but 845 of its 1,171 rows are missing. Would you rank
  it with confidence?

## Challenge 2: Species List (15 min)

- **Base:** `count(species)`: find the same fish spelled differently
  (`"Amercian Eel"`?). Recode with `case_when()`, keep the rest with
  `.default = species`, count again. How many labels remain?
- **Extension:** how many lakes report each species in the 2000s?
  (hint: `n_distinct()`). Are `"Drum"` and `"Freshwater Drum"` the same
  fish? What would you check first?

```{r}
#| echo: false
countdown(minutes = 15)
```

## Break (10 min)

```{r}
#| echo: false
countdown(minutes = 10)
```

## Integrated Challenge (30 min)

**Lake whitefish, Lake Michigan, U.S. catch:** total catch by decade, with
the number of years and missing years beside each total. Which decade
peaked?

- **Base:** one pipeline, filter, `decade` with `mutate()`, `group_by()`,
  `summarize()`, sort. Answer in pounds, not thousands
- **Extension:** the 2010s look low. Compute catch **per recorded year**.
  Does the story change? Which decade "wins" now, and should you trust it?

```{r}
#| echo: false
countdown(minutes = 30)
```

## Every Fix Raises The Next Question

- By total, the **1990s** peak: 69,967 thousand lb (about 70 million lb)
- The 2010s look halved, but the data end in 2015: **6 years**, not 10
- Per year, the 2010s dip only a little (4,265 vs 4,575)
- Per year, the **1870s** "win", on **one** recorded year

. . .

**Report a rate with the number of years behind it.**

## Assignment 05: Park Species Decisions

- **Data:** NPS species lists, 15 most-visited national parks (61k records)
- **Decide:** which records count as "in the park"; one category for two
  kinds of "unknown"
- **Summarize:** percent non-native per park. Watch the **denominator**
- **Investigate:** `dplyr::across()` to total several columns at once
- **Due Sun Oct 11, 11:59 PM.** Last push before then is graded

## How It Is Graded (10 pts)

```{r}
#| echo: false
tibble::tribble(
  ~Criterion, ~Points,
  "Renders + `.html` pushed", 2,
  "Cleaning and types", 2,
  "Core tasks", 3,
  "Investigate", 1,
  "Written answers", 1,
  "README", 1
) |>
  slide_table() |>
  cols_align(align = "center", columns = Points)
```

- Only `assignment/` and `README.md` are checked. **`class/` never is.**
- A red X is feedback, not a grade.

## Before You Leave

- [ ] Follow-along runs top to bottom after restarting R
- [ ] **Commit and push `class/`** (required, not graded; participation).
  GitHub Desktop: summary, **Commit to main**, **Push origin**
- [ ] Push shows on GitHub
- [ ] You know where Assignment 05 lives: `assignment/assignment.qmd`
